```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# multilevelmod

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/multilevelmod)](https://CRAN.R-project.org/package=multilevelmod)
<!-- badges: end -->

multilevelmod enables the use of multi-level models (such as mixed-effects models or Bayesian hierarchical models) with the parsnip package.  

## Installation

You can install the released version of multilevelmod from [CRAN](https://CRAN.R-project.org) with:

``` r
# not yet
install.packages("multilevelmod")
```

For the development version: 

``` r
devtools::install_github("tidymodels/multilevelmod")
```

## Example

Loading mixedlevelmod will trigger it to add a few modeling _engines_ to the parsnip model database. For Bayesian models, there are now `stan-glmer` engines for `linear_reg()`, `logistic_reg()`, and `poisson_reg()`. 

To use these, the function `parsnip::fit()` function should be used instead of `parsnip::fit_xy()` so that the model terms can be specified using the `lme`/`lme4` syntax. 

The `sleepstudy` data is used as an example: 

```{r sleep-lme}
library(multilevelmod)
data(sleepstudy, package = "lme4")

mixed_model_spec <- linear_reg() %>% set_engine("lme4")

mixed_model_fit <- 
  mixed_model_spec %>% 
  fit(Reaction ~ Days + (Days | Subject), data = sleepstudy)

mixed_model_fit
```

For a Bayesian model: 


```{r sleep-stan}
hier_model_spec <- linear_reg() %>% set_engine("stan-glmer")

hier_model_fit <- 
  hier_model_spec %>% 
  fit(Reaction ~ Days + (Days | Subject), data = sleepstudy)

hier_model_fit
```


