---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# multilevelmod

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/multilevelmod)](https://CRAN.R-project.org/package=multilevelmod)
[![Codecov test coverage](https://codecov.io/gh/topepo/multilevelmod/branch/master/graph/badge.svg)](https://codecov.io/gh/topepo/multilevelmod?branch=master)
[![R build status](https://github.com/topepo/multilevelmod/workflows/R-CMD-check/badge.svg)](https://github.com/topepo/multilevelmod/actions)
<!-- badges: end -->

multilevelmod enables the use of multi-level models (such as mixed-effects models or Bayesian hierarchical models) with the parsnip package.  

## Installation

You can install the released version of multilevelmod from [CRAN](https://CRAN.R-project.org) with:

``` r
# not yet
install.packages("multilevelmod")
```

For the development version: 

``` r
devtools::install_github("tidymodels/multilevelmod")
```

## Example

Loading mixedlevelmod will trigger it to add a few modeling _engines_ to the parsnip model database. For Bayesian models, there are now `stan-glmer` engines for `linear_reg()`, `logistic_reg()`, and `poisson_reg()`. 

To use these, the function `parsnip::fit()` function should be used instead of `parsnip::fit_xy()` so that the model terms can be specified using the `lme`/`lme4` syntax. 

The `sleepstudy` data is used as an example: 

```{r sleep-lme}
library(multilevelmod)
data(sleepstudy, package = "lme4")

mixed_model_spec <- linear_reg() %>% set_engine("lmer")

mixed_model_fit <- 
  mixed_model_spec %>% 
  fit(Reaction ~ Days + (Days | Subject), data = sleepstudy)

mixed_model_fit
```

For a Bayesian model: 


```{r sleep-stan}
hier_model_spec <- linear_reg() %>% set_engine("stan-glmer")

hier_model_fit <- 
  hier_model_spec %>% 
  fit(Reaction ~ Days + (Days | Subject), data = sleepstudy)

hier_model_fit
```

## Contributing

This project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

- For questions and discussions about tidymodels packages, modeling, and machine learning, please [post on RStudio Community](https://rstd.io/tidymodels-community).

- If you think you have encountered a bug, please [submit an issue](https://github.com/tidymodels/multilevelmod/issues).

- Either way, learn how to create and share a [reprex](https://rstd.io/reprex) (a minimal, reproducible example), to clearly communicate about your code.

- Check out further details on [contributing guidelines for tidymodels packages](https://www.tidymodels.org/contribute/) and [how to get help](https://www.tidymodels.org/help/).

