---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The multilevelmod package is a [parsnip](https://parsnip.tidymodels.org/) extension package for multi-level models, which are also known as mixed-effects models, Bayesian hierarchical models, etc.
The models wrapped by the multilevelmod package tend to have somewhat different interfaces than the average R modeling package, mostly due to how random effects and independent experimental units are specified. 

This vignette is an overview of how to fit these models. For brevity, we only discuss linear models but the syntax also works for binomial and Poisson outcomes. 

```{r}
#| include: false
library(tidymodels)
library(multilevelmod)
```

```{r}
#| results: hide

library(tidymodels)
library(multilevelmod)

tidymodels_prefer()
theme_set(theme_bw())
```

We'll use some single factor repeated measures experiment data from the lme4 package, on the effect of sleep deprivation on reaction time.  

```{r}
#| fig.alt: > 
#|   18 line plots, one per subject, showing 'Days' on the x-axis and 'Reaction'
#|   on the y-axis. All lines trend upwards but vary across subjects.

data(sleepstudy, package = "lme4")

sleepstudy |> 
  ggplot(aes(x = Days, y = Reaction)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~ Subject) 
```

To show how prediction works, let's create a new data frame for a hypothetical subject "one": 

```{r}
new_subject <- tibble(
  Days = 0:9, 
  Subject = "one"
)
```

Now let's look at how the modeling engines work with multilevelmod. 

## Generalized estimator equations (GEE)

This engine requires the gee package to be installed.

There are no random effects in this model. Like the generalized least squares model discussed below, this model deals with the within-subject correlations by estimating a correlation (or covariance) matrix that is not diagonal. To do this, the model formula should use the `id_var()` function. This is a special syntax for creating model matrices (there is no actual `id_var()` function) that designates the column for the independent experimental unit. 

The correlation structure can be passed as an engine argument: 

```{r}
#| label: gee
gee_spec <- 
  linear_reg() |> 
  set_engine("gee", corstr = "exchangeable")

gee_fit <- 
  gee_spec |> 
  fit(Reaction ~ Days + id_var(Subject), data = sleepstudy)

gee_fit
```

Only a single column name can be given to `id_var()`. 

When predicting, the `id_var` column is not required: 

```{r}
predict(gee_fit, new_subject |> select(Days)) |> 
  bind_cols(new_subject)
```

## Generalized least squares

This engine requires the nlme package to be installed.

For this model, the syntax to specify the independent experimental unit is inside of the `corrrelation` argument for `nlme::gls()`. We'll pass that as an engine argument. Possible values can be found using `?nlme::corStruct`. 

For example:

```{r}
#| label: gls
library(nlme) # <- Only need to load this to get cor*() functions

gls_spec <- 
  linear_reg() |> 
  set_engine("gls", correlation = corCompSymm(form = ~ 1 | Subject))

gls_fit <- 
  gls_spec |> 
  fit(Reaction ~ Days, data = sleepstudy)

gls_fit
```

As with the GEE model, only the regression terms are required for prediction:

```{r}
#| label: gls-pred
predict(gls_fit, new_subject |> select(Days)) |> 
  bind_cols(new_subject)
```

## Linear mixed effects via lme

This engine requires the nlme package to be installed. 

For models created by `nlme::lme()`, the random effects are specified in an argument called `random`. This can be passed via `set_engine()`. The formula specified for `fit()` should only include the fixed effects for the model. 

To fit the basic random intercept model: 

```{r}
#| label: lme
lme_spec <- 
  linear_reg() |> 
  set_engine("lme", random = ~ 1 | Subject)

lme_fit <- 
  lme_spec |> 
  fit(Reaction ~ Days, data = sleepstudy)

lme_fit
```

For predictions, tidymodels uses only the "population effects", i.e., no-subject specific random effects. We have designed tidymodels so that you should not know about the specific training set values when making any type of prediction. 

For lme fit objects, the subject column, if given, is ignored. When the [underlying `predict()` function](https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/predict.lme.html) is used, the `level = 0` argument is automatically invoked: 

```{r}
#| label: lme-pred
predict(lme_fit, new_subject) |> 
  bind_cols(new_subject)

# For this design, this is the same prediction as a training set point:
predict(lme_fit, sleepstudy |> filter(Subject == "308"))
```

## Models using lmer, glmer, and stan_glmer

The `"lmer"`, `"glmer"`, and `"stan_glmer"` engines all use the same formula syntax for fitting multilevel models. See [Section 2.1 of _Linear Mixed Models with lme4_](https://cran.r-project.org/package=lme4/vignettes/lmer.pdf) for details. In this section, we'll demonstrate using the `"lmer"` engine. 

All of the model specification occurs in the formula; no models terms are specified via `set_engine()` (although other arguments can be passed there, as usual). To fit the same random intercept model, the syntax is: 

```{r}
#| label: lmer
lmer_spec <- 
  linear_reg() |> 
  set_engine("lmer")

lmer_fit <- 
  lmer_spec |> 
  fit(Reaction ~ Days + (1|Subject), data = sleepstudy)

lmer_fit
```

We predict in the same way.

```{r}
#| label: lmer-pred
predict(lmer_fit, new_subject) |> 
  bind_cols(new_subject)
```

To determine what packages are required for a model, use this function:

```{r}
required_pkgs(lmer_spec)
```

For the `"stan_glmer"` engine, some relevant arguments that can be passed to `set_engine()` are: 

* `chains`: A positive integer specifying the number of Markov chains. The default is 4.
* `iter`: A positive integer specifying the number of iterations for each chain (including warmup). The default is 2000.
* `seed`: The seed for random number generation. 
* `cores`: Number of cores to use when executing the chains in parallel.
* `prior`: The prior distribution for the (non-hierarchical) regression coefficients. 
* `prior_intercept`: The prior distribution for the intercept (after centering all predictors). 

See `?rstanarm::stan_glmer` and `?rstan::sampling` for more information.

## Using tidymodels workflows

If you use [workflows](https://www.tmwr.org/workflows.html), we have a few suggestions. 

First, instead of using `add_formula()`, we suggest using `add_variables()`. This passes the columns as-is to the model fitting function. To add the random effects formula, use the `formula` argument of `add_model()`. For example: 


```{r}
#| label: wflow
lmer_wflow <- 
  workflow() |> 
  add_variables(outcomes = Reaction, predictors = c(Days, Subject)) |> 
  add_model(lmer_spec, formula = Reaction ~ Days + (1|Subject))

lmer_wflow |> fit(data = sleepstudy)
```

If using a recipe, make sure that functions like `step_dummy()` do **not** convert the column for the independent experimental unit (i.e. subject) to dummy variables. The underlying model fit functions require a single column for these data. 

Using a recipe also offers the opportunity to [set a different role](https://recipes.tidymodels.org/reference/roles.html) for the independent experiment unit, which can come in handy when more complex preprocessing is needed.

```{r}
#| label: rec
rec <- 
  recipe(Reaction ~ Days + Subject, data = sleepstudy) |>
  add_role(Subject, new_role = "exp_unit") |>
  step_zv(all_predictors(), -has_role("exp_unit"))

lmer_wflow |>
  remove_variables() |>
  add_recipe(rec) |>
  fit(data = sleepstudy)
```


### Recipes that rename or replace columns

`step_zv()` leaves the predictor columns as-is; it only ever removes columns that have zero variance. Other steps, such as `step_dummy()` or `step_pca()`, create new columns with different names (or a different number of columns) from the ones in the original data. This matters because the `formula` argument of `add_model()` is evaluated on the data *after* the recipe has been baked, not on the original columns supplied to `recipe()`. If the formula still refers to the original column names, fitting will fail. 

To illustrate, we'll add an artificial factor predictor to `sleepstudy`: 

```{r}
#| label: period
sleepstudy_period <-
  sleepstudy |>
  mutate(period = factor(if_else(Days < 5, "early", "late")))
```

```{r}
#| label: rec-dummy
rec_dummy <-
  recipe(Reaction ~ Days + period + Subject, data = sleepstudy_period) |>
  add_role(Subject, new_role = "exp_unit") |>
  step_dummy(period)
```

`step_dummy()` replaces the `period` column with a new, differently named indicator column. We can see this by prepping and baking the recipe: 

```{r}
#| label: rec-dummy-names
rec_dummy |>
  prep() |>
  bake(new_data = NULL) |>
  names()
```

If the model formula still refers to `period`, the fit will fail because that column no longer exists once the recipe has run: 

```{r}
#| label: rec-dummy-wrong
#| error: true
workflow() |>
  add_recipe(rec_dummy) |>
  add_model(lmer_spec, formula = Reaction ~ Days + period + (1 | Subject)) |>
  fit(data = sleepstudy_period)
```

One fix is to use the column name that the recipe actually produces, `period_late`: 

```{r}
#| label: rec-dummy-right
workflow() |>
  add_recipe(rec_dummy) |>
  add_model(
    lmer_spec,
    formula = Reaction ~ Days + period_late + (1 | Subject)
  ) |>
  fit(data = sleepstudy_period)
```

Naming columns explicitly works, but it is fragile: the formula has to be updated every time the recipe changes, and it breaks down entirely when the *number* of columns a step produces depends on a tuning parameter, such as `step_pca(num_comp = tune())`. There, the number of principal component columns (`PC1`, `PC1 + PC2`, ...) is different for each candidate value in the tuning grid, so no single hand-written list of column names can cover all of them. 

A more robust option is to use `.` in the model formula. Since the formula is evaluated on the baked data, `.` expands to "every remaining column in the baked data except the outcome", whatever those columns happen to be named. Because `.` also picks up the experimental unit column, it needs to be removed again from the fixed effects (it is still used in the random-effects term): 

```{r}
#| label: rec-dummy-dot
workflow() |>
  add_recipe(rec_dummy) |>
  add_model(lmer_spec, formula = Reaction ~ . + (1 | Subject) - Subject) |>
  fit(data = sleepstudy_period)
```

This gives the same fit as the explicit formula above, but it also keeps working when the recipe's output changes, including when tuning `num_comp`: 

```{r}
#| label: rec-pca-tune
data(riesby, package = "multilevelmod")

rec_pca <-
  recipe(depr_score ~ ., data = riesby) |>
  add_role(subject, new_role = "exp_unit") |>
  update_role(week, new_role = "id variable") |>
  step_normalize(all_numeric_predictors(), -has_role("exp_unit")) |>
  step_pca(all_numeric_predictors(), -has_role("exp_unit"), num_comp = tune())

lmer_wflow_pca <-
  workflow() |>
  add_recipe(rec_pca) |>
  add_model(lmer_spec, formula = depr_score ~ . + (1 | subject) - subject)

riesby_split <- group_initial_split(riesby, group = subject)
folds <- group_vfold_cv(training(riesby_split), v = 5, group = subject)

grid_results <- tune_grid(
  lmer_wflow_pca,
  resamples = folds,
  grid = tibble(num_comp = 1:4),
  metrics = metric_set(rmse)
)

collect_metrics(grid_results)
```

No matter how many principal components a given value of `num_comp` produces, `.` picks up all of them, so this one formula is valid across the entire tuning grid.

## Other tips

Finally, there are excellent helper functions in the broom.mixed and tidybayes packages. If these need the underlying model fit object: the `extract_fit_engine()` function can be used on either parsnip or workflow objects: 

```{r}
lmer_wflow |> 
  fit(data = sleepstudy) |> # <- returns a workflow
  extract_fit_engine()       # <- returns the lmer object
```
